Facility and Time Step Discretization Effects in Fuel Cycle Simulation

### 2015-09-29 paper.tex
more edits

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of a few different modeling
choices that may sometimes be taken for granted in the fuel cycle modeling
domain.  Four scenarios were compared using combinations of either fleet-based
or individual reactors with either monthly or quarterly (3-month) time steps.
When under supply-constraint pressure, the four cases exhibit noticably
different behavior.  Otherwise, they generally behaved very similarly.
Fleet-based modeling is more efficient in supply-constrained environments at
the expense of losing insight on issues such as realistic sub-optimal fuel
distribution and challenges in reactor refueling cycle staggering.
Finer-grained time steps also enable more efficient material use in
supply-constrained environments.  Large simulations with fleet-based reactors
run much more quickly than their individual reactor counterparts.

@@

@@

    \item groups of facilities (usually all facilities of a given type) are
        treated as a single, aggregate entity.

    \item resource flows are often (approximately) continuous (limited by time
        step discretization)

@@

Many fuel cycle simulators step through time in discrete steps.  The duration
of a single time step can have significant impact on the dynamics of a
simulation.  The time step duration can affect the magnitude of individual
material transfers between facilities and correspondingly affects the noise
level in facility material inventories and flows.  It can also act as a
minimum bound for outage durations. The effects of the time step duration will
also be investigated together with fleet vs facility modeling.

@@

Cyclus was designed to operate natively in unix/linux type environments with
flexible and scriptable usage and input file format.  Cyclus holds its diverse
generated data into a powerful open-source database format.  In addition to
information about material flows and facility deployments, Cyclus' single-file
database contains version information about the archetypes used in a
simulation, the simulation input file, version of Cyclus used to run the
simulation, and the installed versions of Cyclus dependencies. This and other
features of Cyclus lend themselves well to automation, research provenance and
reproducibility, and large-scale computing.

@@

The scenario used is a transition from a the current once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle known as
EG-23 in the DOE's Advanced Nuclear Fuel Cycle Options report
\cite{wigeland_nuclear_2014}. Scenario and transition details were patterned
after ongoing work by the DOE in their Fuel Cycle Options campaign.

The scenario starts with 100 LWRs and follows an exponential curve with a 1\%
annual growth rate for 200 years.  The decommissioning of the initial 100
reactors is staggered over years 15 to 55.  Enrichment feeds LWRs with fresh
fuel with infinite capacity. In year 35, fast reactors become available for
deployment and no more thermal reactors are built.  LWR spent fuel separations
begins in year 15 with 2000 MTHM/yr capacity and increases to 3000 MTHM/yr in
year 25.  Fast reactor spent fuel separations and fuel fabrication have
infinite capacity.  All spent fuel is stored/cooled for 7 years before it is
available for separations. Figure \ref{fig:flow} shows the material flow
relationships between the facilities.  Compositions used for the simulations
were simple, containing primarily $^{238}U$, $^{235}U$, $^{239}Pu$,
$^{241}Am$, and small quantities of a few lanthanides.

@@

Cyclus version ?.? (TODO:~update~version) was used for this analysis.
Facility models used were the standard ones provided with Cyclus in the
Cycamore library (also version ?.?).  A custom storage facility was used
because Cycamore does not yet provide one.  The standard cycamore reactor was
used for both reactor types (LWR and SFR) in cases 1 and 3, and the custom
fleet-based reactor described earlier also used in cases 2 and 4.
Additional parameters for the reactor types are shown in Table
\ref{tab:reactors}.

@@

\begin{table}[h]
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
            & LWR & FR \\
        \hline                       
        Lifetime (yr)         & 80 & 80 \\
        Cycle length (months) & 18 & 14 \\
        Batch size (kg)       & 29565 & 7490 \\
        Batches per core      & 3 & 5 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \caption[Reactor facility parameters]{Reactor facility parameters. The cycle length includes the refueling outage.}
    \label{tab:reactors}
\end{table}

The described scenario is used as closely as possible in each of the four
cases.  However, a few other small case-dependent changes to the scenario
relating to reactor configuration are necessary.  For the cases using
quarterly time steps with individual reactors, cycle and outage times must be
adjusted to be a multiple of 3 months.  In order to make the comparison more
fair, the cases with monthly time steps are also adjusted to have identical
cycle and outage times.  The invariants preserved with respect to reactor
behavior are shown in Table \ref{tab:invar}.  Table \ref{tab:reactors} shows
the computed/selected configuration for both reactor types for all four cases.

>>>>

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of a few different modeling
choices that may sometimes be taken for granted in the fuel cycle modeling
domain.  Four scenarios are compared using combinations of either fleet-based
or individual reactors with either monthly or quarterly (3-month) time steps.
The Cyclus fuel cycle simulator is used as a level playing field to compare
the scenarios.  When under supply-constraint pressure, the four cases exhibit
noticably different behavior.  Otherwise, they generally behaved very
similarly.  Fleet-based modeling is more efficient in supply-constrained
environments at the expense of losing insight on issues such as realistically
sub-optimal fuel distribution and challenges in reactor refueling cycle
staggering.  Finer-grained time steps also enable more efficient material use
in supply-constrained environments.  Large simulations with fleet-based
reactors run much more quickly than their individual reactor counterparts.

@@

\pagebreak

@@

    \item Groups of facilities (usually all facilities of a given type) are
        treated as a single, aggregate entity.

    \item Resource flows are often (approximately) continuous (limited by time
        step discretization).

@@

Many fuel cycle simulators step through time in discrete steps.  The duration
of a single time step can have significant impact on the dynamics of a
simulation.  The time step duration can affect the magnitude of individual
material transfers between facilities and correspondingly affects the noise
level in facility material inventories and flows.  It can also act as a
minimum bound for outage durations. The effects of the time step duration are
also investigated together with fleet vs facility modeling.

@@

Cyclus was designed to operate natively in unix/linux type environments with
flexible and scriptable usage and input file format.  Cyclus stores its
diverse generated data in a powerful open-source database format.  In addition
to information about material flows and facility deployments, Cyclus'
single-file database contains the simulation input file, version information
about the archetypes used in a simulation, the version of Cyclus used to run
the simulation, and the installed versions of Cyclus dependencies. This and
other features of Cyclus lend themselves well to automation, research
provenance and reproducibility, and large-scale computing.

@@

The scenario used approximates a transition from a the current once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle known as
EG-23 in the DOE's Advanced Nuclear Fuel Cycle Options report
\cite{wigeland_nuclear_2014}. Scenario and transition details were patterned
after ongoing work by the DOE in their Fuel Cycle Options campaign.

The scenario starts with 100 LWRs and follows an exponential curve with a 1\%
annual growth rate for 200 years.  The decommissioning of the initial 100
reactors is staggered over years 15 to 55.  Enrichment feeds LWRs with fresh
fuel with infinite capacity. In year 35, fast reactors become available for
deployment and no more thermal reactors are built.  LWR spent fuel separations
begins in year 15 with 2000 MTHM/yr capacity and increases to 3000 MTHM/yr in
year 25.  Fast reactor spent fuel separations and fuel fabrication have
infinite capacity.  All spent fuel is stored/cooled for 7 years before it is
available for separation. Figure \ref{fig:flow} shows the material flow
relationships between the facilities.  Compositions used for the simulations
were simple containing primarily $^{238}U$, $^{235}U$, $^{239}Pu$, $^{241}Am$,
and small quantities of a few lanthanides.  Full composition details can be
found in the Cyclus input files used \cite{Carlsen2015}.

@@

A pre-release of Cyclus version 1.4 was used for this analysis.  Archetypes
used were the standard ones provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage facility was
used because Cycamore did not yet provide one.  The standard cycamore reactor
was used for both LWR and SRF reactor types in cases 1 and 3.  Correspondingly
the custom fleet-based reactor described earlier was used for both reactor
types in cases 2 and 4.  Additional configuration for the reactor types are
shown in Table \ref{tab:reactors}.

@@

\begin{table}[h]
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
            & LWR & SFR \\
        \hline                       
        Lifetime (yr)         & 80 & 80 \\
        Cycle length (months) & 18 & 14 \\
        Batch size (kg)       & 29565 & 7490 \\
        Batches per core      & 3 & 5 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \caption[Reactor facility parameters]{Reactor facility parameters. Cycle length includes the refueling outage.}
    \label{tab:reactors}
\end{table}

The described scenario is used as closely as possible in each of the four
cases.  A few other small case-dependent changes to the scenario relating to
reactor configuration are necessary.  For the cases using quarterly time steps
with individual reactors, cycle and outage times must be adjusted to be a
multiple of 3 months.  In order to make the comparison more fair, the cases
with monthly time steps are also adjusted to have identical cycle and outage
times.  The invariants preserved with respect to reactor behavior between all
four cases are shown in Table \ref{tab:invar}.  Table \ref{tab:reactors} shows
the computed/selected configuration for both LWR and SFR reactor types for all
four cases.
